---
output:
  pdf_document: default
  html_document: default
---
# Functions and Graphs

Reading Materials: \newline
Croft, A. and R. Davidson *Foundation maths.* (Harlow: Pearson, 2016) 6th edition. **Chapter 17 Graphs of Functions.**

Larson, R. Precalc with limits. (Boston, MA: Cengage, 2017) 4th edition. **Chapter 1 Functions and their graphs (sections 1.1–1.8).**

## Numbers and Intervals

### Number Sets

Mathematics often begins by classifying **numbers** into different sets. Each set expands on the previous one, allowing us to solve more and more problems.

**1. Natural Numbers ($\mathbb{N}$)**

* The counting numbers:

  \begin{equation}
  \mathbb{N} = \{1, 2, 3, 4, \dots \}
  \end{equation}
* Sometimes defined to include $0$:

  \begin{equation}
  \mathbb{N}_0 = \{0, 1, 2, 3, \dots \}
  \end{equation}
* Example: number of students in a class, number of apples.

*Whether natural number includes zero, depends on context and who you ask. For the sake of clarity, use $\mathbb{N}_0$ when it includes zero.*

**2. Integers ($\mathbb{Z}$)**

* Extends natural numbers by including **zero** and **negative numbers**:

  \begin{equation}
  \mathbb{Z} = \{\dots, -3, -2, -1, 0, 1, 2, 3, \dots\}
  \end{equation}
* Example: temperature in Celsius, bank account balance (can be negative).

**3. Rational Numbers ($\mathbb{Q}$)**

* Numbers that can be written as a **fraction** of two integers:

  \begin{equation}
  \mathbb{Q} = \left\{ \frac{p}{q} : p, q \in \mathbb{Z}, q \neq 0 \right\}
  \end{equation}
* Includes integers (since $5 = \tfrac{5}{1}$).
* Decimal expansions of rationals are **terminating** (e.g. $0.75$) or **repeating** (e.g. $0.333...$).
* Example: $\tfrac{2}{3}, -\tfrac{7}{5}, 4.25$.

**4. Real Numbers ($\mathbb{R}$)**

* All numbers that can be placed on a **continuous number line**.
* Includes both **rational** and **irrational numbers**.

  * **Irrational**: cannot be written as a fraction, decimal expansion is infinite and non-repeating.

    * Examples: $\pi, \sqrt{2}, e$.
* Example: every measurement of length, mass, or time in physics is modeled as a real number.

**5. Complex Numbers ($\mathbb{C}$)**

* Extends real numbers by introducing the **imaginary unit** $i$, where $i^2 = -1$.
* General form:

  \begin{equation}
  z = a + bi \quad (a,b \in \mathbb{R})
  \end{equation}
* Real part = $a$, Imaginary part = $b$.
* Example: $3 + 4i$.
* Complex numbers are essential in advanced math, physics, and engineering (e.g. AC circuits, quantum mechanics).

**6. Irrational Number**

* The set of **irrational numbers** is

  \begin{equation}
  \mathbb{R} \setminus \mathbb{Q}
  \end{equation}

  i.e. all real numbers that are **not** rational. Examples: $\pi, \sqrt{2}, e$.

* Every real number is **either rational or irrational**, so we can write:

  \begin{equation}
  \mathbb{R} = \mathbb{Q} \cup \left(\mathbb{R} \setminus \mathbb{Q}\right)
  \end{equation}

* These two subsets are **disjoint**:

  \begin{equation}
  \mathbb{Q} \cap \left(\mathbb{R} \setminus \mathbb{Q}\right) = \varnothing
  \end{equation}

\begin{equation}
\mathbb{N} \subset \mathbb{Z} \subset \mathbb{Q} \subset \mathbb{R} \subset \mathbb{C}
\end{equation}

### Intervals on the Real Line

An **interval** is a set of real numbers between two endpoints $a$ and $b$ ($a < b$).

There are different types depending on whether endpoints are included or excluded:

**1. Closed Interval**

$$
[a, b] = \{x \in \mathbb{R} : a \leq x \leq b\}
$$

* Both $a$ and $b$ are included.
* Example: $[2,5] = \{x: 2 \leq x \leq 5\}$.
<center>
```{r, echo=FALSE, fig.width=3, fig.height=2}
library("intervals")
# x is a collection of intervals
# that are all left-closed and right-open
x = Intervals(
  matrix(c(2, 5), nrow = 1, byrow = TRUE),
  closed = c(TRUE, TRUE), # intervals are of the form [a, b)
  type = "Z" # inputs are integers
)
plot(x, xlim = c(0, 6))
```
</center>

**2. Open Interval**

$$
(a, b) = \{x \in \mathbb{R} : a < x < b\}
$$

* Neither $a$ nor $b$ is included.
* Example: $(2,5) = \{x: 2 < x < 5\}$.

<center>
```{r, echo=FALSE, fig.width=3, fig.height=2}
library("intervals")
# x is a collection of intervals
# that are all left-closed and right-open
x = Intervals(
  matrix(c(2, 5), nrow = 1, byrow = TRUE),
  closed = c(FALSE, FALSE), # intervals are of the form [a, b)
  type = "Z" # inputs are integers
)
plot(x, xlim = c(0, 6))
```
</center>


**3. Half-Open Intervals**

* Left-closed, right-open:

  $$
  [a, b) = \{x \in \mathbb{R} : a \leq x < b\}
  $$
  <center>
```{r, echo=FALSE, fig.width=3, fig.height=2}
library("intervals")
# x is a collection of intervals
# that are all left-closed and right-open
x = Intervals(
  matrix(c(2, 5), nrow = 1, byrow = TRUE),
  closed = c(TRUE, FALSE), # intervals are of the form [a, b)
  type = "Z" # inputs are integers
)
plot(x, xlim = c(0, 6))
```
</center>
  
* Left-open, right-closed:

  $$
  (a, b] = \{x \in \mathbb{R} : a < x \leq b\}
  $$
<center>
```{r, echo=FALSE, fig.width=3, fig.height=2}
library("intervals")
# x is a collection of intervals
# that are all left-closed and right-open
x = Intervals(
  matrix(c(2, 5), nrow = 1, byrow = TRUE),
  closed = c(FALSE, TRUE), # intervals are of the form [a, b)
  type = "Z" # inputs are integers
)
plot(x, xlim = c(0, 6))
```
</center>

**4. Infinite Intervals**

* $(a, \infty) = \{x \in \mathbb{R} : x > a\}$
* $[a, \infty) = \{x \in \mathbb{R} : x \geq a\}$
* $(-\infty, b) = \{x \in \mathbb{R} : x < b\}$
* $(-\infty, b] = \{x \in \mathbb{R} : x \leq b\}$

Note: $\infty$ and $-\infty$ are **not real numbers**, so intervals involving them are always open on that side.

<center>
```{r, echo=FALSE, fig.width=3, fig.height=3}
library("intervals")
# x is a collection of intervals
# that are all left-closed and right-open
x = Intervals(
  matrix(c(-Inf, 5,
           2, Inf), nrow = 2, byrow = TRUE),
  closed = c(TRUE, TRUE), # intervals are of the form [a, b)
  type = "Z" # inputs are integers
)
plot(x, xlim = c(0, 6))
```
</center>

## Functions

### Introduction to functions

### Type of Functions

### Composite Functions

### Inverse Functions

### Properties of Function

## Graph Sketching

## Application of Functions